CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-12 codecheck-paper.tex
adjust figure style to rest, add terms
restructure small bits and slightly shorten conclusions

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/rr.pdf}
  \caption{The inverse problem in reproducible research.  The left
  half of the diagram shows the diverse range of materials used
  within a laboratory.  These materials are often then
  condensed for sharing with the outside world via the
  research paper, a static PDF document.  Working backwards from the
  PDF to the underlying materials is impossible. This prohibits reuse
  and is not only non-transparent for a specific paper, but also 
  ineffective for science as a whole. By sharing the
  materials on the left, others outside the lab can reproduce
  or build on this work.}
  \label{fig:inverse}
\end{figure}

@@

CODECHECK works!
We have created a considerable number of certificates to demonstrate it.
The creation of the certificates and the interactions with authors and
editors not only improved the principles and the workflow, but also
confirmed the approach taken. This result corroborates findings from
similar evaluations of reproducible computational research in journals and
at conferences.
CODECHECKs and can increase trust in scientific results and the set of 
shared principles and name can through recognition value allow researchers
across publication venues to judge the level of scrutiny that results have
faced. CODECHECK requires direct acknowledgement of the contribution to the
scientific community in the role, not indirectly via citations of 
reproductions, and can be a worthwhile educational activity, especially for
ECRs. A technically simple yet far reaching change would be a special type of 
reviewer activity for codechecking in public researcher profiles next to
conventional peer reviews.

Yet, CODECHECK underlies the same limitations as peer review
in general and is closely connected to larger disruptions and challenges
in scholarly communication
\cite{eglen_recent_2018,tennant_ten_2019,fyfe_mission_2019}, 
including the tensions between commercial publishing and reviewer's often
free labour, and a global pandemic jumbling up scientific publishing
and exposing a broader general audience to science, even preliminary 
results \cite{munafo_what_2020}.
Even more, it is clear that establishing CODECHECK-like systems can be 
impeded by and must be seen in connection to much larger issues in 
science, such as broken metrics or malpractices triggered by publication
pressure \cite{piwowar_altmetrics:_2013,nosek_promoting_2015}.
We certainly do not want the binary attribute of "CODECHECK completed"
to become a factor in bibliometric approaches for performance assesments.
% Mittermaier, Peer Review and Bibliometrics: http://hdl.handle.net/2128/22745
While developed for the current "paper"-centric publication process and
being compatible with different styles of peer review, the CODECHECK 
principles would also work well with novel publication paradigms, e.g.,
peer-reviewed computational notebooks \cite{earthcube_new_2020},
iterative and granular communication of scientific outputs,
such as Octopus\footnote{\url{https://science-octopus.org/}}, 
articles with live-code \cite{perkel_pioneering_2019-1}
\footnote{For example, eLife's ERA, see 
\url{https://elifesciences.org/labs/dc5acbde/welcome-to-a-new-era-of-reproducible-publishing}.},
decentralized infrastructure and public reviewer reputation systems
\cite{tenorio-fornes_towards_2019},
and completely new visions for scholarly communication and peer review
\footnote{Cf., for example, \emph{A modern vision for peer review} by 
Amy J. Ko: \url{https://medium.com/bits-and-behavior/a-modern-vision-for-peer-review-d5f73f0fae07}}.
An explicit segmentation of research steps could even make the focus 
of a CODECHECK easier by only checking the ``analysis'' sub-publication.
The discovery of CODECHECKs could be increased by depositing the reports
into public databases of reproductions, such as \emph{SciGen.Report}.
Public researcher profiles, such as ORCID, may consider different
types of reviewer activity to capture the contributions of independent 
code execution to science.
Notably, the discussed limitations are largely self-imposed for easier
acceptance and evolutionary integration, as to not break the current 
system and increase demands over time without leaving practitioners behind.

If adopted today, a CODECHECK system, even if temporal in a sustainable 
transition towards more open publication and review practices, can contribute
to increased trust in the results of research. Introducing CODECHECK
should be informed by lessons learned from (introducing) open peer review
(OPR) \cite{ross-hellauer_guidelines_2019}.
All our conversations with publishers and editors at some point
touched on the procedural and technical implications of openness, be it 
that OPR is neither applied nor well-known by stakeholders, or that process
management software would not support such open communication.
We would be keen to use the flexibility of the principles and cooperate
on non-public CODECHECKs to learn more about the advantages and yet unclear
specific challenges. Does CODECHECK work better with OPR?
Each rollout of CODECHECK should be accompanied by studies to ensure 
codechecking does not fall into the same gaps as peer review did \cite{tennant_limitations_2020}
and to ensure positive change within the respective peer review system.
More reproducible practices initiated by CODECHECKs could lead
communities/disciplines to reach a state where authors do provide good
enough material and reviewers have acquired broady enough skills that
the latter will generally conduct a CODECHECK-level of checking, and
only in especially sophisticated cases will a specialised codechecker
be included.
The main challenge for us remains to get journals to embrace the
idea behind CODECHECK and realise processes conform to the principles,
whether they use the actual name for it or not.
The CODECHECK badge is, intentionally, not branded beyond the checkmark
and green colour and simply states ``code works''.
This \emph{cultural change}, however, is needed for the valuation of the
efforts that go into proper evaluation of papers.
Multiple adopting journals will also help to answer relevant questions on
the establishing process (What are crucial decisions or pain points? Can
authors retract code/data once a CODECHECK has started?),
what variants of CODECHECKs will be most common, and how open CODECHECKs
may influence the scope and anonymity of conventional review over time.

With the implementation of CODECHECK workflows, the question of training
codecheckers will become relevant. We expect both a mentoring scheme within
the CODECHECK community (experienced codecheckers provide on the job training
or serve as fallback advisors) as well as collaboration with reproducible
research initiatives such as
ReproHack\footnote{\url{https://reprohack.github.io/reprohack-hq/}},
ReproducibiliTea\footnote{\url{https://reproducibilitea.org/}}
\cite{fitzgibbon_brewing_2020},
and Repro4Everyone\footnote{\url{https://repro4everyone.org/}}
\cite{auer_reproducibility_2020},
to be able to grow the pool of codecheckers as needed.
The initial reaction of the scientific community, where up to today over 
twenty people spontaneously signed up as volunteer codecheckers after talks
on the initiative, makes us feel optimistic that the time is right for 
putting scholarly peer review on the path to facilitate sharing and execution
of computer programs and documenting reproducibility of results.

>>>>

\begin{figure}
% emojis: https://github.com/googlefonts/noto-emoji
% cross mark: u274c
% check mark: u2714
  \centering
  \includegraphics[width=0.7\textwidth]{figs/rr.pdf}
  \caption{The inverse problem in reproducible research.  The left
  half of the diagram shows the diverse range of materials used
  within a laboratory.  These materials are often then
  condensed for sharing with the outside world via the
  research paper, a static PDF document.  Working backwards from the
  PDF to the underlying materials is impossible. This prohibits reuse
  and is not only non-transparent for a specific paper, but also 
  ineffective for science as a whole. By sharing the
  materials on the left, others outside the lab can reproduce
  or build on this work.}
  \label{fig:inverse}
\end{figure}

@@

CODECHECK works!
We have created a considerable number of certificates to demonstrate it.
The creation of the certificates and the interactions with authors and
editors not only improved the principles and the workflow, but also
confirmed the approach taken. This result corroborates findings from
similar evaluations of reproducible computational research in journals and
at conferences.
CODECHECKs can increase trust in scientific results.
The set of 
shared principles and common name, through recognition value,
allow researchers
across publication venues to judge the level of scrutiny that results have
faced. CODECHECK requires direct acknowledgement of the contribution to the
scientific community in the role of codechecker, not indirectly via citations 
of reproductions or informal crediation and closed inner circles.
Codechecking can be a worthwhile educational activity, especially for
ECRs.

Yet, CODECHECK underlies the same limitations as peer review
in general and is closely connected to larger disruptions and challenges
in scholarly communication
\cite{eglen_recent_2018,tennant_ten_2019,fyfe_mission_2019}, 
including the tensions between commercial publishing and reviewer's often
free labour, and a global pandemic jumbling up scientific publishing
and exposing a broader general audience to science, even preliminary 
results \cite{munafo_what_2020}.
It is clear that establishing CODECHECK-like systems can be 
impeded by and must be seen in connection to much larger issues in 
science, such as broken metrics or malpractices triggered by publication
pressure \cite{piwowar_altmetrics:_2013,nosek_promoting_2015}.
We certainly do not want the binary attribute of ``code works''
to become a factor in bibliometric approaches for performance assesments.
% Mittermaier, Peer Review and Bibliometrics: http://hdl.handle.net/2128/22745

While developed for the current "paper"-centric publication process and
being compatible with different styles of peer review, the CODECHECK 
principles would also work well with novel publication paradigms, e.g.,
peer-reviewed computational notebooks \cite{earthcube_new_2020},
iterative and granular communication of scientific outputs, 
articles with live-code \cite{perkel_pioneering_2019-1}
\footnote{For example, eLife's ERA, see 
\url{https://elifesciences.org/labs/dc5acbde/welcome-to-a-new-era-of-reproducible-publishing}.},
decentralized infrastructure and public reviewer reputation systems
\cite{tenorio-fornes_towards_2019},
and completely new visions for scholarly communication and peer 
review\footnote{Cf., for example, \emph{A modern vision for peer review} by 
Amy J. Ko: \url{https://medium.com/bits-and-behavior/a-modern-vision-for-peer-review-d5f73f0fae07}}.
An explicit segmentation of research steps could even make the focus 
of a CODECHECK easier by only checking the ``analysis'' sub-publication.
The discovery of CODECHECKs could be increased by depositing the reports
into public databases of reproductions, such as \emph{SciGen.Report}.
Public researcher profiles, such as ORCID, may consider different
types of reviewer activity to capture the contributions of independent 
code execution to science.
Notably, the discussed limitations are largely self-imposed for easier
acceptance and evolutionary integration, as to not break the current 
system and increase demands over time without leaving practitioners behind.

If adopted today, a CODECHECK system, even if temporal in a sustainable 
transition towards more open publication and review practices, can contribute
to increased trust in the results of research. Introducing CODECHECK
should be informed by lessons learned from (introducing) open peer review
(OPR) \cite{ross-hellauer_guidelines_2019}.
All our conversations with publishers and editors at some point
touched on the procedural and technical implications of openness, be it 
that OPR is neither applied nor well-known by stakeholders, or that process
management software would not support such open communication.

More reproducible practices initiated by CODECHECKs could lead
communities/disciplines to reach a state where authors do provide good
enough material and reviewers have acquired broady enough skills that
the latter will generally conduct a CODECHECK-level of checking, and
only in especially sophisticated cases will a specialised codechecker
be included.
The main challenge for us remains to get journals to embrace the
idea behind CODECHECK and realise processes conform to the principles,
whether they use the actual name for it or not.
We would be keen to use the flexibility of the principles and cooperate
with journals to learn more about the advantages and yet unclear
specific challenges -- Do CODECHECKs really work better with OPR?
Each rollout of CODECHECK should be accompanied by studies to ensure 
codechecking does not fall into the same gaps as peer review did
\cite{tennant_limitations_2020}
and to ensure positive change within the respective peer review system.
The CODECHECK badge is, intentionally, not branded beyond the checkmark
and green colour and simply states ``code works''.
This \emph{cultural change}, however, is needed for the valuation of the
efforts that go into proper evaluation of papers.
Multiple adopting journals will also help to answer relevant questions on
the establishing process (What are crucial decisions or pain points? Can
authors retract code/data once a CODECHECK has started?),
what variants of CODECHECKs will be most common, and how open CODECHECKs
may influence the scope and anonymity of conventional review over time.

With the implementation of CODECHECK workflows, the question of training
codecheckers will become relevant. We expect both a mentoring scheme within
the CODECHECK community (experienced codecheckers provide on the job training
or serve as fallback advisors) as well as collaboration with reproducible
research initiatives such as
ReproHack\footnote{\url{https://reprohack.github.io/reprohack-hq/}},
ReproducibiliTea\footnote{\url{https://reproducibilitea.org/}}
\cite{fitzgibbon_brewing_2020},
and Repro4Everyone\footnote{\url{https://repro4everyone.org/}}
\cite{auer_reproducibility_2020},
to be able to grow the pool of codecheckers as needed.
The initial reaction of the scientific community and the spontaneously signing
up volunteer codecheckers make us feel optimistic that the time is right for 
putting scholarly peer review on the path to facilitate sharing and execution
of computational workflows.
